Splitting of the homology of the punctured mapping class group

### 2018-10-29 HBraidSurf.tex
Finished proving the main theorem!

\section{Homology of configuration spaces of surfaces}
We want to understand the homology groups $H_*(\bms)=H_*(C_m(\sg))$: we
are interested in these groups as they appear as the homology of the fiber
in the Leray-Serre spectral sequence associated to
the Birman exact sequence \ref{eq:Birman} or, which is the same, to the bundle
\ref{eq:Birmanbundle}. In particular we are interested in $H_*(C_m(\S))$
\emph{as a $\Z_2-$representation of the group $\gg$.}
% From now on $\Z_2-$coefficients in homology
% and cohomology will always be understood.

In \cite{LM} L\"offler and Milgram proved that this must be a symplectic
representation of the mapping class group. By \emph{symplectic} we mean the following:
\begin{defn}
 \label{defn:symplrep}
 Let $\H=H_1(\S)\simeq\Z_2^{2g}$.
 The natural action of $\gg$ on $\H$ induces a surjective map
 $\gg\to Sp_{2g}(\Z_2)$. A representation of $\gg$ over $\Z_2$ is called \emph{symplectic}
 if it is a pull-back of a representation of $Sp_{2g}(\Z_2)$ along this map.
\end{defn}

In \cite{BCM} B\"odigheimer, Cohen and Taylor computed $H_*(C_m(\sg))$ \emph{as a graded $\Z_2-$vector space}.
Their method provides all Betti numbers, but the action of $\gg$ cannot be easily deduced, as
their descripition of $H_*(C_m(\sg))$ depends on a handle decomposition of $\sg$, which is not preserved,
not even up to isotopy, by diffeomorphisms of $\sg$.

In this section and in the next one we will prove the following theorem; to the best of the author's knowledge
it doesn't appear in the literature.
\begin{thm}
 \label{thm:Hbms*as*ggrep}
 There is an isomorphism of bigraded $\Z_2-$representations of $\gg$
 \[
  \bigoplus_{m\geq 0} H_*\pa{C_m(\S)}\simeq \Z_2\left[Q^j\epsilon\,|\, j\geq 0\right]\otimes\Sym_{\bullet}(\H).
 \]
 Here we mean the following:
 \begin{itemize}
  \item the bidegree is given on left by homological degree $*$ and by the direct summand,
  indexed by $m$,
  on which the homology class is supported, i.e. by the number $m$ of points
  involved in constructing the homology class; we call $*$ the \emph{degree} and $m$ the \emph{weight},
  and write $(*,m)$ for the bidegree;
  \item for $j\geq 0$, $Q^j\epsilon$ is the image in $H_{2^j-1}(C_{2^j}(\S))$ of a generator
  of the group $H_{2^j-1}(C_{2^j}(\Sigma_{0,1}))\simeq \Z_2$ 
  under the natural map induced by an embedding $\Sigma_{0,1}\hookrightarrow\S$,
  and $\Z_2\left[Q^j\epsilon\,|\, j\geq 0\right]$ is the polynomial ring on
  infinitely many variables $\epsilon,Q\epsilon,Q^2\epsilon,\dots$;
  \item $\H=H_1(\sg)$ is identified with $H_1(C_1(\S))$ in a natural way, and $\Sym_{\bullet}(\H)$ is the
  symmetric algebra on $\H$;
  \item degrees and weights are extended on right by the usual multiplicativity-additivity rule;
  \item the action of $\gg$ on right is the diagonal action on the product: it is a trivial action
  on the factor $\Z_2[Q^j\epsilon\,|\, j\geq 0]$ and it is the symplectic action induced by the action on $\H$ on the
  factor $\Sym_{\bullet}(\H)$.
  \end{itemize}

  Notice that for any bi-homogeneus element in the right-hand side, the weight is greater or equal than
  the degree: indeed factors of the form $Q^j\epsilon$ have weight strictly higher than the degree,
  whereas factors in $\H$ and its symmetric powers have equal weight and degree.
\end{thm}
In this section we will prove that there is an isomorphism of \emph{bigraded $\Z_2-$vector spaces}
as in theorem \ref{thm:Hbms*as*ggrep}; in the next section we will deal with the action of
$\gg$. In these two sections we abbreviate $\S=\sg$.

Since we work with $\Z_2-$coefficients, it is equivalent to compute homology or cohomology,
and in this section we will prefer to compute, for each bidegree $(*,m)$, $H^*(\cms)$ as a graded
$\Z_2-$vector space; the dual graded vector space $H_*(\cms)$ is of course abstractly isomorphic,
i.e. it has the same dimension.

We will mimic the method used by Fuchs (\cite{Fuchs:CohomBraidModtwo}) to compute the $\Z_2-$cohomology
of the configuration space of the plane.
As already said this computation recovers a known result, but it has the advantage of
being quite elementary and of providing a part of
the geometric insight that we will need in the next section.

@@

We first introduce a space $\T(\S)$ which is homeomorphic to $\mrS$, the interior of $\S$. The construction
corresponds to a handle decomposition of $\S$ with one $0-$handle and $2g$ $1-$handles.
\begin{defn}
\label{defn:Tsg}
If $g=0$, hence $\S=\sg$ is the disc, we set $\T(\S)=(0,1)^2$, the interior of the unit square. Assume
now $g\geq 1$.

@@

For $1\leq i\leq g$, we identify $I^l_{2i-1}$ with $I^r_{2i}$, obtaining an open interval $\U_i\subset\T(\S)$,
and we identify $I^r_{2i-1}$ with $I^l_{2i}$, obtaining an open interval $\V_i\subset\T(\S)$.
Each interval $\U_i,\V_i$ has a natural parametrisation by $(0,1)$.
\end{defn}
The space $\T(\S)$ is homeomorphic to $\mrS$. From now
on we will identify the two open surfaces and in particular we will identify $\cms$ with the
space of configurations of $m$ points in $\T(\S)$.

@@

Indeed in $F_p/F_{p-1}$ all outer
differentials vanish (see lemma \ref{lem:doperatoropenmodtwo}
for the definition), in particular the numbers $u_i,v_i$
don't change along the boundary map of $F_p/F_{p-1}$, and this latter chain complex splits as a direct sum
of chain complexes indexed by splittings $m-p=\sum_{i=1}^g (u_i+v_i)$ as above.
It is then immediate to identify the inner differentials
with the ones one would have in the case $g=0$.

We notice that $\tCh_*(C_p(\Sigma_{0,1}))$ is exactly the chain complex described by Fuchs in
\cite{Fuchs:CohomBraidModtwo}; in particular its homology is generated by symmetric chains:
\begin{defn}
\label{defn:symchain}
Given a splitting $p=\sum_{j=0}^{\infty}\alpha_j2^j$ of $p$ into powers of 2, with multiplicities $\alpha_i$,
the associated symmetric chain in $\tCh_*(C_p(\Sigma_{0,1}))$
is the sum of all $e^{\tup}$ where $\tup$ ranges among all tuples
$\pa{l,(x_i)_{i\leq l}}$ such that
\begin{itemize}
 \item $l=\sum_{i=1}^{\infty}\alpha_i$;
 \item every $x_i$ is a power of 2;
 \item for all $j\geq 0$ there are exactly $\alpha_j$ indices $i$ such that $x_i=2^j$.
\end{itemize}
Here and in the following it is understood that $\alpha_j=0$ for all but finitely many indices $j$.
\end{defn}

If we run the Leray spectral sequence associated to the filtered chain complex $\tCh_*(\cms)$,
the $E^1-$page contains on the $p-$th column the homology of $F_p/F_{p-1}$; as we have seen this consists of
many copies of the homology of $\tCh_*(C_p(\Sigma_{0,1}))$, each generated by symmetric chains.

@@

In particular the spectral sequence collapses on its first page and $H_*(\cms^c)$
has a $\Z_2$-basis given by generalised symmetric chains. We have the following lemma:
\begin{lem}
\label{lem:gensymchain}
The homology $H_*(\cms^c)$ has a graded basis given by the classes $[p,(\alpha_j),(u_i,v_i)]$
associated to generalised symmetric chains.
% this basis is bijection with
% the choices of the following set of data:
% \begin{itemize}
%  \item a number $0\leq p\leq m$;
%  \item a splitting $m-p=\sum_{i=1}^g (u_i+v_i)$ with $u_i,v_i\geq 0$;
%  \item a splitting $p=\sum_{j=0}^{\infty}\alpha_j2^j$ with $\alpha_j\geq 0$.
% \end{itemize}
% The homology class $[p,(u_i,v_i),(\alpha_j)]\in H_*(\cms^c)$ has homological degree
% $m+\sum_{j=1}^{\infty}\alpha_i$.
\end{lem}
% To interpret this class
% as a monomial in the tensor product of theorem \ref{thm:Hbms*as*ggrep} we need the following definition.

@@

\begin{defn}
 \label{defn:dualHbasis}
We can see the $\U_i$'s and $\V_i$'s as properly embedaded $1-$manifolds in $\mrS$;
by Poincaré-Lefschetz duality they represent classes in $H_1\pa{(\mrS)^c}\simeq H^1(\S)$,
and in particular they form a basis
of this cohomology group. We call $\u_i,\v_i\in H_1\pa{\mrS}$ the dual basis.

>>>>

\section{Homology of configuration spaces of surfaces}
\label{sec:HBraidSurf}
We want to understand the homology groups $H_*(\bms)=H_*(C_m(\sg))$: we
are interested in these groups as they appear as the homology of the fiber
in the Leray-Serre spectral sequence associated to
\ref{eq:Birman}, \ref{eq:Birmanbundle} or \ref{eq:BirmanbundleD}.
In particular we are interested in $H_*(C_m(\S))$
as a $\Z_2-$representation of the group $\gg$.

In \cite{LM} L\"offler and Milgram proved that this must be a $\Z_2-$symplectic
representation of the mapping class group. By \emph{$\Z_2-$symplectic} we mean the following:
\begin{defn}
 \label{defn:symplrep}
 Let $\H=H_1(\S)\simeq\Z_2^{2g}$.
 The natural action of $\gg$ on $\H$ induces a surjective map
 $\gg\to Sp_{2g}(\Z_2)$. A representation of $\gg$ over $\Z_2$ is called \emph{$\Z_2-$symplectic}
 if it is a pull-back of a representation of $Sp_{2g}(\Z_2)$ along this map.
\end{defn}

In \cite{BCT} B\"odigheimer, Cohen and Taylor computed $H_*(C_m(\sg))$ \emph{as a graded $\Z_2-$vector space}.
Their method provides all Betti numbers, but the action of $\gg$ cannot be easily deduced, as
their descripition of $H_*(C_m(\sg))$ depends on a handle decomposition of $\sg$, which is not preserved,
not even up to isotopy, by diffeomorphisms of $\sg$.

In this section and in the next one we will prove the following theorem; to the best of the author's knowledge
it doesn't appear in the literature.
\begin{thm}
 \label{thm:Hbms*as*ggrep}
 There is an isomorphism of bigraded $\Z_2-$representations of $\gg$
 \[
  \bigoplus_{m\geq 0} H_*\pa{C_m(\sg)}\simeq \Z_2\left[Q^j\epsilon\,|\, j\geq 0\right]\otimes\Sym_{\bullet}(\H).
 \]
 Here we mean the following:
 \begin{itemize}
  \item the bidegree is given on left by homological degree $*$ and by the direct summand,
  indexed by $m$,
  on which the homology class is supported, i.e. by the number $m$ of points
  involved in constructing the homology class; we call $*$ the \emph{degree} and $m$ the \emph{weight},
  and write $(*,m)$ for the bidegree;
  \item for $j\geq 0$, $Q^j\epsilon$ is the image in $H_{2^j-1}(C_{2^j}(\sg))$ of a generator
  of the group $H_{2^j-1}(C_{2^j}(\D))\simeq \Z_2$ 
  under the natural map induced by the embedding $\D\hookrightarrow\mrS$,
  and $\Z_2\left[Q^j\epsilon\,|\, j\geq 0\right]$ is the polynomial ring on
  infinitely many variables $\epsilon,Q\epsilon,Q^2\epsilon,\dots$;
  \item $\H=H_1(\sg)$ is identified with $H_1(C_1(\S))$ in a natural way, and $\Sym_{\bullet}(\H)$ is the
  symmetric algebra on $\H$;
  %\item degrees and weights are extended on right by the usual multiplicativity-additivity rule;
  \item the action of $\gg$ on right is the diagonal action on the product: it is a trivial action
  on the factor $\Z_2[Q^j\epsilon\,|\, j\geq 0]$ and it is the $\Z_2-$symplectic action on the
  factor $\Sym_{\bullet}(\H)$ induced by the $\Z_2-$symplectic action on $\H$.
  \end{itemize}
\end{thm}
Notice that for any bi-homogeneus element in the right-hand side, the weight is greater or equal than
the degree: indeed factors of the form $Q^j\epsilon$ have weight strictly higher than the degree,
whereas factors in $\H$ and its symmetric powers have equal weight and degree.

Notice that in the case $g=0$ the group $\Gamma_{0,1}$ is trivial and the previous theorem
reduces to equation \ref{eq:Cohen}.
% isomorphism of bigraded vector spaces
% \begin{equation}
% \label{eq:genuszero}
% \bigoplus_{m\geq 0} H_*\pa{C_m(\Sigma_{0,1})}\simeq \Z_2[Q^j\epsilon\,|\, j\geq 0].
% \end{equation}
% This isomorphism was essentially proved by Fuchs in \ref{Fuchs:CohomBraidModtwo}. Moreover the left hand
% side is the homology of the space $\coprod_{m\geq 0} C_m(\Sigma_{0,1})$, which has a natural structure
% of algebra over the operad $E^2$ of little $2-$cubes. The isomorphism in equation \ref{eq:genuszero}
% is an isomorphism of rings (using the Pontryagin product on the left hand side), and $Q$ represents here
% the Araki-Kudo-Dyer-Lashof operation $Q\colon H_*(X;\Z_2)\to H_{2*+1;\Z_2}(X)$,
% which is defined for any $E^2-$algebra $X$; $\epsilon\in H_0(C_1(\Sigma_{0,1}))$ is the generator.
% This explains our notation; we refer to Cohen \cite[Chap.3]{CLM} for more details.

In this section we will prove that there is an isomorphism of \emph{bigraded $\Z_2-$vector spaces}
as in theorem \ref{thm:Hbms*as*ggrep}; in the next section we will deal with the action of
$\gg$.
%In these two sections we abbreviate $\S=\sg$.

Since we work with coefficients in the field $\Z_2$, it is equivalent to compute homology or cohomology,
and in this section we will prefer to compute $H^*(\cms)$ for all bidegrees $(*,m)$.
% the dual graded vector space $H_*(\cms)$ is abstractly isomorphic,
% i.e. it has the same dimension.

We will mimic the method used by Fuchs (\cite{Fuchs:CohomBraidModtwo}) to compute the $\Z_2-$cohomology
of $C_m(\D)$.
As already said this computation recovers a known result, but it has the advantage of
being quite elementary and of providing a part of
the geometric insight that we will need in the next section.

@@

We first introduce a space $\T(\S)$ which is homeomorphic to $\mrS$, the interior of $\S$. The construction
corresponds to a handle decomposition of $\S$ with one $0-$handle and $2g$ $1-$handles.

\begin{defn}
\label{defn:Tsg}
If $g=0$, hence $\S=\sg$ is the disc, we set $\T(\S)=(0,1)^2$, the interior of the unit square. Assume
now $g\geq 1$.

@@

For $1\leq i\leq g$, we identify $I^l_{2i-1}$ with $I^r_{2i}$, obtaining an open interval $\U_i\subset\T(\S)$,
and we identify $I^r_{2i-1}$ with $I^l_{2i}$, obtaining an open interval $\V_i\subset\T(\S)$.
Each interval $\U_i,\V_i$ has a natural parametrisation by $(0,1)$.

The space $\T(\S)$ is homeomorphic to $\mrS$. From now
on we will identify the two open surfaces and in particular we will identify $\cms$ with the
space of configurations of $m$ points in $\T(\S)$.

Let $\D\subset\mrS$ be the open square $(1/4;3/4)\times(1/2,1)$:
$\D$ is an open disc in $\mrS$ near $\partial\S$ and 
it is disjoint from all $\U_i$'s and $\V_i$'s.
\end{defn}

@@

Indeed in $F_p/F_{p-1}$ all outer
differentials vanish (see lemma \ref{lem:doperatoropenmodtwo}
for the definition), in particular the numbers $u_i,v_i$
don't change along the boundary map of $F_p/F_{p-1}$, and this latter chain complex splits as a direct sum
of chain complexes indexed by splittings $m-p=\sum_{i=1}^g (u_i+v_i)$ as above.
It is then immediate to identify the inner differentials
with the ones one would have in the case $g=0$, i.e. for the surface $\Sigma_{0,1}$.

We notice that $\tCh_*(C_p(\Sigma_{0,1})^c)$ is exactly the chain complex described by Fuchs in
\cite{Fuchs:CohomBraidModtwo}, so we pause for a moment and
recall some classical facts about the cohomology of configuration
spaces of the disc.
\begin{defn}
\label{defn:symchain}
Given a splitting $p=\sum_{j=0}^{\infty}\alpha_j2^j$ of $p$ into powers of 2, with multiplicities $\alpha_j$,
the associated symmetric chain in $\tCh_*(C_p(\Sigma_{0,1})^c)$
is the sum of all $e^{\tup}$ where $\tup$ ranges among all tuples
$\pa{l,(x_i)_{i\leq l}}$ such that
\begin{itemize}
 \item $l=\sum_{j=1}^{\infty}\alpha_j$;
 \item every $x_i$ is a power of 2;
 \item for all $j\geq 0$ there are exactly $\alpha_j$ indices $i$ such that $x_i=2^j$.
\end{itemize}
Here and in the following it is understood that $\alpha_j=0$ for all but finitely many indices $j$.

We denote by $|(\alpha_j)_{j\geq 0}|\in\tCh_{p+l}(C_p(\Sigma_{0,1})^c)$ the symmetric chain
and by $[(\alpha_j)_{j\geq 0}]\in\tH_{p+l}(C_p(\Sigma_{0,1})^c)$ the associated homology class.
\end{defn}
In \cite{Fuchs:CohomBraidModtwo} Fuchs shows that a graded basis for $\tH_*(C_p(\Sigma_{0,1})^c)$
is given by the collection of all classes $[(\alpha_j)_{j\geq 0}]$ satisfying $\sum\alpha_j2^j$.

Using Poincaré-Lefschetz duality we obtain a basis for the cohomology of $C_p(\Sigma_{0,1})$;
the dual basis of $H_*(C_p(\Sigma_{0,1}))$
happens to be the basis of monomials
$\prod_{j\geq 0}(Q^i\epsilon)^{\alpha_j}\in H_*(C_p(\Sigma_{0,1}))$ of weight $m$, using
the isomorphism \ref{eq:Cohen} in its full meaning (i.e. as an isomorphism of rings,
where $Q$ denotes the first Dyer-Lashof operation).

% This notation is compatible, after interpreting equation \ref{eq:genuszero} as an isomorphism
% of rings, with the results of \cite[Chap.3]{CLM}: the monomials constructed there, using
% the Pontryagin product and the operation $Q$, are really the dual basis of the basis of
% symmetric chains for cohomology.

We will not need this finer result in this article, so the expression
$\prod_{j\geq 0}(Q^i\epsilon)^{\alpha_j}\in H_{p-l}(C_p(\Sigma_{0,1}))$ will only denote
the (unique) homology class whose algebraic intersection with $[(\alpha'_j)]\in\tH_{p+l}(C_p(\Sigma_{0,1})^c)$
is $1\in\Z_2$ if and only if the equality $\alpha_j=\alpha'_j$ holds for all $j\geq 0$.
% i.e. we \emph{define} monomials as the dual basis of the basis of symmetric chains, and keep considering
% \ref{eq:Cohen} only as an isomorphism of bigraded vector spaces.

We now go back to the filtered chain complex $\tCh_*(\cms^c)$:
if we run the associated Leray spectral sequence,
the $E^1-$page contains on the $p-$th column the homology of $F_p/F_{p-1}$; as we have seen this consists of
many copies of the homology of $\tCh_*(C_p(\Sigma_{0,1}))$, each generated by symmetric chains.

@@

In particular the spectral sequence collapses on its first page and
% $H_*(\cms^c)$
% has a $\Z_2$-basis given by generalised symmetric chains.
we have the following lemma:
\begin{lem}
\label{lem:gensymchain}
The homology $H_*(\cms^c)$ has a graded basis given by the classes $[p,(\alpha_j),(u_i,v_i)]$
associated to generalised symmetric chains.
% this basis is bijection with
% the choices of the following set of data:
% \begin{itemize}
%  \item a number $0\leq p\leq m$;
%  \item a splitting $m-p=\sum_{i=1}^g (u_i+v_i)$ with $u_i,v_i\geq 0$;
%  \item a splitting $p=\sum_{j=0}^{\infty}\alpha_j2^j$ with $\alpha_j\geq 0$.
% \end{itemize}
% The homology class $[p,(u_i,v_i),(\alpha_j)]\in H_*(\cms^c)$ has homological degree
% $m+\sum_{j=1}^{\infty}\alpha_j$.
\end{lem}
% To interpret this class
% as a monomial in the tensor product of theorem \ref{thm:Hbms*as*ggrep} we need the following definition.

@@

\begin{defn}
 \label{defn:dualHbasis}
We can see the $\U_i$'s and $\V_i$'s as properly embedaded $1-$manifolds in $\mrS$;
by Poincaré-Lefschetz duality they represent classes in $H_1\pa{(\mrS)^c}\simeq H^1(\S)$,
and in particular they form a basis
of the latter cohomology group. We call $\u_i,\v_i\in H_1\pa{\mrS}$ the dual basis.

### 2018-10-29 Introduction.tex
Finished proving the  gg-equivariant isomorphism
Finished proving the main theorem!

The associated Leray-Serre spectral sequence in $\Z_2-$homology has a second page $E^2_{p,q}=H_p(\gg;H_q(\bms;\Z_2))$,
and converges to $H_{p+q}(\ggm;\Z_2)$.

Our main theorem is that the spectral sequence has already collapsed in its second page.
\begin{thm}
\label{thm:main}
For each $l\geq 0$ there is an isomorphism
\begin{equation}
\label{eq:main}
 H_l(\ggm;\Z_2)\cong \bigoplus_{p+q=l} H_p(\gg;H_q(\bms;\Z_2)).
\end{equation}
\end{thm}
Thus the computation of $H_*(\ggm;\Z_2)$ reduces to the computation of the homology of $\gg$ with
twisted coefficients in the representation $H_*(\bms;\Z_2)$: as we will see the $\ggm-$representation
$H_*(\bms;\Z_2)$ splits as a direct sum of symmetric powers of $H_1(\sg)$.
In particular this representation is symplectic: this had already benn stated in \cite{LM}.

@@

From now on $\Z_2$-coefficients will always be understood if no coefficients are specified.

>>>>

The associated Leray-Serre spectral sequence $E(m)$ in $\Z_2-$homology has a second page $E(m)^2_{k,q}=H_k(\gg;H_q(\bms;\Z_2))$,
and converges to $H_{k+q}(\ggm;\Z_2)$.

Our main theorem is that the spectral sequence has already collapsed in its second page.
\begin{thm}
\label{thm:main}
For all $l\geq 0$ there is an isomorphism
\begin{equation}
\label{eq:main}
 H_l(\ggm;\Z_2)\cong \bigoplus_{k+q=l} H_k(\gg;H_q(\bms;\Z_2)).
\end{equation}
\end{thm}
Thus the computation of $H_*(\ggm;\Z_2)$ reduces to the computation of the homology of $\gg$ with
twisted coefficients in the representation $H_*(\bms;\Z_2)$: as we will see the $\ggm-$representation
$H_*(\bms;\Z_2)$ splits as a direct sum of symmetric powers of $H_1(\sg)$.
In particular this representation is symplectic: this had already benn stated in \cite{LM}.

@@
